Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-07-12 manuscript/manuscript.qmd
van hees spt chatter

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

In an effort to mature the methods for estimating sleep from thigh-worn
accelerometers, we evaluated various models for predicting in-bed and
sleep time and their derived sleep quality parameters. Furthermore, we
trained and evaluated the models using raw and median-filtered
gold-standard predictions from the ZM. In general, all sequential models
performed well at predicting in-bed time. More challenging was it to
distinguish wake from sleep on the extracted in-bed time, however, the
performance of the sequential models were enhanced by the application of
median filters. Moreover, even though the multiclass biLSTM showed good
performance across F1 score, precision and NPV, the derived sleep
quality parameters were not on par with the XGBoost model which
demonstrated the highest performance metrics across all evaluations,
including epoch-to-epoch prediction and sleep quality parameter
derivatives. Despite this, all sequential models showed low specificity
values, indicating difficulty in correctly classifying awake epochs
during time in bed. The application of 5-minute and 10-minute median
filters improved the performance metrics of all models, with the XGBoost
model consistently leading. The filters increased total sleep time and
sleep efficiency, while reducing wake after sleep onset and the number
of awakenings. The XGBoost model also showed the least bias and highest
correlation with ZM across sleep quality parameters.

@@

To the best of our knowledge, Johansson and
colleagues[@johansson_development_2023] are the only researchers who
have reported epoch-to-epoch performance metrics for sleep scoring using
thigh-worn accelerometers, beyond just "waking time" and "in-bed time."
They achieved a mean sensitivity of 0.84, specificity of 0.55, and
accuracy of 0.80, using a single-night evaluation dataset of 71
subjects. Despite our models achieving a sensitivity above 97%, they,
like Johansson et al.'s algorithm, struggled with detecting in-bed awake
epochs. This is reflected in the low specificity scores, ranging from
54.7% to 76.4%, reported in our study. The challenge of low specificity
is not unique to methods using thigh-worn devices. Conley et al.'s
meta-analysis[@conley2019] reported similar findings when estimating
sleep using wrist-worn accelerometers among healthy adults, with a mean
sensitivity, accuracy, and specificity of 0.89, 0.88, and 0.53,
respectively. These figures align with our results regarding in-bed
time. Patterson and colleagues[@patterson_40_2023] recently summarized
the performance of various heuristic algorithms, machine learning, and
deep learning models used to predict sleep. They found the mean
sensitivity and specificity to be 93% (SD = 2.8) and 60% (SD = 11.1)
respectively. These findings underscore the challenge of automating the
detection of in-bed awake periods. Interestingly, despite low
specificity values for most of our models and configurations, we
observed an overestimation of LPS and WASO, contrasting with most
previous research[@conley2019; @palotti2019]. This overestimation of
wake epochs is evident from the low NPV scores, indicating that only a
small proportion of the wake predictions are actually correct. This
discrepancy may be driven by the SMOTE process used to balance the
dataset. If the synthetic "wake" examples created by SMOTE are not
representative of the true "wake" data, the models might learn to
incorrectly classify certain "sleep" epochs as "wake". This could lead
to an overestimation of LPS and WASO, as the models are incorrectly
identifying more periods of wakefulness during the sleep period.

often times imbalanced datasets only containing night recordings

The sleep quality parameter that our models consistently corresponded
most with the ZM derived sleep quality parameters were SPT and TST

something on outliers...maybe related to incl 180 degrees problem.

single night vs multiple nights/days

It is unclear whether Johansson et al.[@johansson_development_2023]\]
provide a method for detecting the in-bed time which hinders its
applicability for researchers that want to ...

A significant finding was the impact of median filters on the sleep
metrics. Implementing 5-minute and 10-minute median filters led to
notable changes in key sleep parameters. In particular, the Total Sleep
Time (TST), Sleep Efficiency (SE), and Latency to Persistent Sleep (LPS)
increased, suggesting that the filters may categorize some instances of
wakefulness as sleep and smooth out brief awakenings. This underscores
the potential of median filters in refining sleep data and mitigating
the potential misclassification of sleep states.

Interestingly, the most marked change was observed in Wake After Sleep
Onset (WASO), with a substantial reduction in the filtered data. This
reduction, coupled with the decrease in the number of awakenings,
implies a significant role of median filters in sleep-wake cycle
characterization, potentially enhancing the reliability of sleep
assessment.

The machine learning models demonstrated high overall performance on an
epoch-to-epoch basis, with the XGBoost model marginally outperforming
the others. However, the models had more varied performance in
classifying sleep conditions, as evidenced by the sleep performance
metrics. The relatively lower Specificity values across all models
suggest a common challenge in correctly classifying negative instances.

In the precision-recall and ROC curve analyses, the Decision Tree model
showed superior precision-recall AUC, indicating strong predictive
accuracy. In contrast, the XGBoost model consistently demonstrated a
high ROC AUC, indicating a robust ability to differentiate between
classes. However, it's worth noting that the Neural Network model
generally showed weaker performance in these areas.

The three-class biLSTM multiclassifier showed overall good performance,
further supporting the potential of machine learning in sleep study.
However, the absence of Specificity values limits a complete assessment
of its performance, calling for further exploration in future studies.

strengths and limitations

ZM as gold-standard? Without a gold standard, both methods can both
contribute to disagreement.

@@

In conclusion, this study provides compelling evidence of the utility of
machine learning models and median filters in enhancing the precision
and reliability of pediatric sleep assessment. However, more research is
needed to overcome the challenge of correctly classifying negative
instances and to fully understand the performance of these models across
all classes and conditions. The findings underscore the potential of
machine learning in advancing sleep medicine, paving the way for more
accurate, reliable, and personalized sleep assessment in pediatric
populations.

>>>>

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    pdf-engine: lualatex
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  simple-article-pdf:
    indent: false
    classoption: 
     - 9pt
     - lucida
     - twocolumn
     - twoside
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

In an effort to mature the methods for estimating sleep from thigh-worn
accelerometers, we evaluated various models for predicting in-bed and
sleep time and their derived sleep quality parameters. Furthermore, we
trained and evaluated the models using raw and median-filtered gold
standard predictions from the ZM. In general, all sequential models
performed well at predicting in-bed time. More challenging was it to
distinguish wake from sleep on the extracted in-bed time, however, the
performance of the sequential models were enhanced by the application of
median filters. Moreover, even though the multiclass biLSTM showed good
performance across F1 score, precision and NPV, the derived sleep
quality parameters were not on par with the XGBoost model which
demonstrated the highest performance metrics across all evaluations,
including epoch-to-epoch prediction and sleep quality parameter
derivatives. Despite this, all sequential models showed low specificity
values, indicating difficulty in correctly classifying awake epochs
during time in bed. The application of 5-minute and 10-minute median
filters improved the performance metrics of all models, with the XGBoost
model consistently leading. The median filters increased total sleep
time and sleep efficiency, while reducing wake after sleep onset and the
number of awakenings. The XGBoost model also showed the least bias and
highest correlation with ZM across sleep quality parameters.

@@

To the best of our knowledge, Johansson and
colleagues[@johansson_development_2023] are the only researchers who
have reported epoch-to-epoch performance metrics for sleep scoring using
thigh-worn accelerometers, beyond just "waking time" and "in-bed time."
They achieved a mean sensitivity of 0.84, specificity of 0.55, and
accuracy of 0.80, using a single-night evaluation dataset of 71
subjects. Despite our models achieving a sensitivity above 97%, they,
like Johansson et al.'s algorithm, struggled with detecting in-bed awake
epochs. This is reflected in the low specificity scores, ranging from
54.7% to 76.4%, reported in our study. The challenge of low specificity
is not unique to methods using data collected from thigh-worn devices.
Conley et al.'s meta-analysis[@conley2019] reported similar findings
when estimating sleep using wrist-worn accelerometers among healthy
adults, with a mean sensitivity, accuracy, and specificity of 0.89,
0.88, and 0.53, respectively. Furthermore, Patterson and
colleagues[@patterson_40_2023] recently summarized the performance of
various heuristic algorithms, machine learning, and deep learning models
used to predict sleep. They found the mean sensitivity and specificity
to be 93% (SD = 2.8) and 60% (SD = 11.1) respectively. These findings
underscore the challenge of automating the detection of in-bed awake
periods. Interestingly, despite low specificity values for most of our
models and configurations, we observed an overestimation of LPS and
WASO, contrasting with most previous research[@conley2019;
@palotti2019]. This overestimation of wake epochs is evident from the
low NPV scores, indicating that only a small proportion of the wake
predictions are actually correct. This discrepancy may be driven by the
SMOTE process used to balance the dataset. If the synthetic "wake"
samples created by SMOTE are not representative of the true "wake" data,
the models might learn to incorrectly classify certain "sleep" epochs as
"wake". This could lead to an overestimation of LPS and WASO, as the
models are incorrectly identifying more periods of wakefulness during
the sleep period.

The use of the SMOTE technique likely improved the performance of our
models by addressing the class imbalance in our data. However, this
technique also introduced synthetic"wake" samples that may not be fully
representative of true wake data. This could potentially lead some
models to overestimate the wake class. Interestingly, the biLSTM model,
which was not trained on SMOTE-processed data, was the only one to
overestimate TST and SE. On the other hand, the XGBoost model, which was
trained on data subjected to the SMOTE process, was able to handle the
synthetic "wake" samples better than the other models, and it did not
overestimate TST to the same degree. The Bland-Altman statistics for the
XGBoost model trained on the 5-minute median filtered ZM predictions
showed a mean difference of -7 minutes for TST and -1.1% for SE, with
limits of agreement ranging from -95.5 to 81.4 minutes and from -15.6%
to 13.3% respectively. This suggests that the XGBoost model was able to
maintain a balance between sensitivity and specificity, and it was not
overly influenced by the synthetic "wake" samples. The XGBoost model's
success with the SMOTE dataset may be due to its ability to handle
non-representative synthetic samples. XGBoost's gradient boosting
mechanism allows it to iteratively learn from the errors of previous
models, which can help it to better distinguish between true wake data
and synthetic wake samples created by SMOTE. This iterative learning
process could make XGBoost more robust to the inaccuracies introduced by
the synthetic samples, leading to better overall performance.

In our evaluation of sleep quality parameters, we found that Latency to
Persistent Sleep (LPS) had the largest mean error relative to absolute
time allocated to LPS. This suggests that the initial epochs of Sleep
Period Time (SPT) are particularly challenging to classify correctly.
This is also supported by the poor Pearson correlations between LPS
derived from model predictions and the ZM. The XGBoost model, which was
the best performer among all models, overestimated LPS by an average of
26.4 minutes for models trained on raw ZM predictions, 28.5 minutes for
models trained on 5-minute filtered ZM predictions, and 34.5 minutes for
models trained on 10-minute filtered ZM predictions. This level of
discrepancy is comparable to the mean error of sleep latency of 23
minutes reported by Johansson et al.[@johansson_development_2023].
Johansson et al. suggest that the discrepancy with the gold standard is
likely due to the multifaceted nature of the sleep state, which is a
complex physiological process. Short awakenings or sleep episodes may
not necessarily correspond to noticeable changes in thigh movement,
making them difficult to detect and accurately classify. These results
align with several methods for wrist-worn devices reviewed by Conley and
colleagues[@conley2019]. They reported correlations between
accelerometer and PSG sleep onset latency (equivalent to LPS) from 10
studies with a mean correlation of 0.2 (ranging from -0.69 to 0.69),
indicating the inherent difficulty in estimating this parameter using
accelerometry alone.

In this study, we used the ZM as the reference method, rather than PSG,
which is considered the gold standard for sleep measurement. This choice
may contribute to discrepancies between our models and the ZM, as
without a true gold standard, it's difficult to determine the source of
disagreement. However, we believe that the use of ZM, which allows for
multiple consecutive nights of recording, is valuable. This approach
captures intra-individual variances in sleep, which is impractical with
PSG. It also enabled us to include more nights in our study typically
compared to those relying on PSG. For instance, the widely used
Newcastle dataset[@hees2015] only contains data from 28 participants.
However, upon examining the ZM outputs, we found that the raw
predictions were not optimal for developing machine learning models due
to a seemingly low signal-to-noise ratio (see @fig-zm-median). The ZM
itself mitigates this issue by applying certain filtering processes when
generating sleep quality parameters. For example, epochs contributing to
WASO must be in contiguous epochs of 3, and sleep only counts towards
sleep quality parameters if 10 out of 12 minutes are scored as sleep. To
improve the prospect of our machine learning algorithms, we applied
median filters to the ZM raw predictions. This did in fact alter the
derived sleep quality parameters. Notably, the mean WASO decreased from
39 minutes in the raw predictions to 30.6 minutes in the 5-minute median
filtered predictions, and further decreased to 22.3 minutes in the
10-minute filtered predictions. The application of 5-minute and
10-minute median filters also led to increases in TST, SE, and LPS. This
suggests that the filters may categorize some instances of wakefulness
as sleep and smooth out brief awakenings. Despite these changes, the
overall sleep quality profile derived from the median-filtered
predictions is still comparable to that from the raw predictions,
justifying our approach.

Our study's XGBoost model demonstrated relatively narrower limits of
agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
LOAs between our XGBoost model and the algorithm developed for
thigh-worn devices by Johansson et al. study[@johansson2022], our
XGBoost model showed narrower LOAs for TST , SE, LPS , and WASO, but not
SPT. Generally, all methods, both from this study and from the reviewed
literature, exibit wide LOAs sugesting that there is high variability in
the derevide sleep quality parameters. These findings imply that the
current methods, are only reasonbly reliable for assessing sleep quality
parameters at a group level. However, caution should be exercised when
applying the models and methods to individual-level sleep assessments.
Therefore, further improvements and refinements are needed to enhance
the precision and reliability of these models for individual sleep
assessments.

Typically, sleep detection methods are applied in two contexts: either
to night recordings or to 24-hour recordings. In night recordings, it is
possible to derive sleep quality parameters like SE and LPS because the
SPT is already known because it is inferred from the length of the
recording[@conley2019; @patterson_40_2023]. On the other hand, when
sleep detection methods are applied to 24-hour recordings, most methods
do not have the ability to infer the SPT with sleep
diaries[@girschik2012]. Consequently, these methods are unable to
generate certain sleep quality parameters that rely on the
SPT[@doherty2017; @anderson2014]. We have attempted to circumvent this
issue by employing models able to distinguish between in-bed awake time
and in-bed asleep time over a 24h recording making our models capable of
estimating all commonly used sleep quality parameters. Van Hees et
al.[@van_hees_estimating_2018] have developed an algorithm to identify
SPT from wrist-worn devices which was recently validated by Plekhanova
and colleagues[@plekhanova2023]. which in combination with other
algorithms or models on said SPT can then be used infer further sleep
quality parameters.

Strengths and limitations It is not always clear the methods for
detecting sleep from accelerometer data is

It is unclear whether Johansson et al.[@johansson_development_2023]
provide a method for detecting the in-bed time which hinders its
applicability for researchers that want to ...

@@

In conclusion, our study has advanced the methods for estimating sleep
from thigh-worn accelerometers by evaluating various models for
predicting in-bed and sleep time. Our results indicate that all
sequential models and the multiclass model performed well at predicting
in-bed time. However, all models showed difficulty in correctly
classifying awake epochs during time in bed, highlighting the challenge
of automating the detection of in-bed awake periods. The application of
median filters improved the performance metrics of all models,
suggesting that this approach may be beneficial in future studies.
Despite these promising results, our findings also underscore the need
for caution when applying these models to individual-level sleep
assessments due to the high variability in derived sleep quality
parameters. Future research should focus on refining these models to
enhance their precision and reliability for individual sleep
assessments.

Our open-source algorithm and the inferred sleep parameters will open
the door to future studies on sleep and sleep architecture using
large-scale accelerometer databases.
